### 2019-11-26 paper.rmd
add rough draft

# Background

@@

## Overview of Methodologies

At present, there are a number of approaches to carrying out person-oriented
analysis when the indicators, of variables, used to find the groups are
continuous. This section is focused upon two approaches: cluster analysis and
Latent Profile Analysis (LPA). We focus on these because they are relatively
simple and accessible forms of person-oriented analysis that can also
accommodate cross sectional and longitudinal data alike. Just as a brief point
of contrast, when the indicators for the groups are dichotomous (or categorical
and can therefore be “dummy-coded” to be dichotomous), an approach closely
related to LPA, Latent Class Analysis, can be used (cite PSU folks).

## Cluster Analysis

Cluster analysis is an approach that has been frequently applied to analysis of
topics in educational psychology (such as motivation and engagement) from a more
holistic, whole-person approach. One particular way in which cluster analysis
has been used has been as part of an approach termed I-States-as-Objects
Analysis (ISOA). The purpose of ISOA is to identify profiles at specific time
points, and then to explore how individuals do or do not shift between profiles
across time points (needs cite). In their early work, Bergman and Magnusson
(1999) and colleagues (1999) described a two-step approach to cluster analysis
as part of their I-States-as-Objects analysis. In this approach, all of the
observations are clustered. The key idea behind the approach is that each
observation represents not a permanent group, but rather a status, or an
i-state. These i-states, are considered independently for the purposes of
constructing profiles. Then, shifts between these i-states across time (say, two
or three time points) are evaluated.

Commonly, cluster analysis carried out as part of ISOA in fact involves two
distinct clustering steps. This is because different clustering algorithms have
different strengths--as well as weaknesses. In ISOA, the first step is Ward’s
hierarchical cluster analysis (Bergman & El-Khouri, 1999). In this step,
observations are split, step-by-step, based upon their multivariate similarity.
The analyst provides the number of groups which will be used and interpreted,
though technically most algorithms will continue splitting the groups until each
observation makes up its own. This type of clustering is deterministic: The same
results will be obtained each time it is run. In the second step of the cluster
analysis, an iterative, k-means cluster analysis is carried out (Bergman and
El-Khouri, 1999). For this step, the algorithm partitions the observations into
groups, the success of the clustering based on a criterion such as the
variability in the data explained by the groups, and then “moves,” when
possible, the observations that are not in the group they are most similar to to
the one that they are most similar to. When no more such moves are possible, the
algorithm stops. On its own, k-means is not deterministic: Depending on the
starting values, it can obtain different results the different times that it is
run. So, the results from the hierarchical cluster analysis, the first step, are
used as start values to initialize the k-means clustering.

The rationale behind using these two approaches is that the hierarchical
clustering provides reasonable guesses for the profile membership, and then the
iterative clustering optimizes the clustering solution. To sum up, then, cluster
analysis as part of ISOA in fact involves two separate cluster analyses. The
merits of this approach are to leverage the benefits of each: the iterative
clustering starts with reasonable cluster solutions generated from the
hierarchical clustering and optimizes them.

## Latent Profile Analysis

Another approach what is becoming increasingly common for carrying out
person-oriented analysis is Latent Profile Analysis (LPA). LPA is a type of
finite mixture model (Harring and Hodis, 2017). The goal of this approach is
estimate the parameters for a number of distributions (often multivariate) from
a single dataset. The approach can be thought of as seeking an answer to the
question, Do the data in a sample come from more than one population? Thus, such
an approach is model-based, and some descriptions in the literature refer to it
as model-based clustering (Hennig, Meila, Murtagh, & Rocci, 2015; Scrucca, Fop,
Murphy, & Raftery, 2017). Thus, one distinction between LPA and cluster analytic
approaches is that LPA is model-based; instead of using algorithms to group
together cases, LPA seeks to estimate parameters - in terms of variances and
covariances and how they are the same or different across profiles - that best
characterize the different distributions. Then, this approach seeks to assign to
each observation a probability that the observation is a sample from the
population associated with each profile (or mixture component).

Because LPA is model-based, a number of different model parameterizations can be
estimated. These models differ in terms of whether--and how--parameters are
estimated across the profiles. These parameters are the means for the different
profiles, which, in this approach, always are estimated freely across the
profiles; the variances for the variables used to create the profiles, which can
be estimated freely or can be estimated to be the same, or equal, across
profiles; and the covariances of the variables used to create the profiles,
which can be freely-estimated, estimated to be equal, or fixed to be zero. One
challenge facing the analyst using LPA is that these parameters and the distinct
model parameterizations that can be estimated is the different terminology used.
As one example, Scrucca et al. (2017) refer to these parameterizations not in
terms of whether and how parameters are estimated, but rather in terms of the
geometric properties of the distributions that result from particular
parameterizations. Muthen and Muthen (1997-2017) and others (Pastor et
al., 2007) commonly refer to local independence to mean that the covariances are
fixed to zero (also described as the specification of the covariance matrix as
“diagonal,” because only the diagonal components, or the variances, are
estimated). More information on the model parameterizations are discussed in the
context of the software tool tidyLPA that we have developed.

This section outlines several LPA models and guidelines/examples for when they
might be most appropriate in person-centered analysis. In general, as more
parameters are estimated (i.e., those that are fixed to zero are estimated as
being equal across profiles; or those estimated as being equal across profiles
are freely-estimated across them), the model becomes more complex; the model may
fit better, but also be overfit, meaning that the profiles identified may be
challenging to replicate with another, separate data set. Even still,
flexibility in terms of which models can be estimated also has affordances. For
example, the varying means, equal variances, and covariances fixed to 0. A
researcher might choose this model specification if she wants to model the
variables to be used to create profiles that are independent. This model is very
simple, as no covariances are estimated and the variances are estimated to be
the same across profiles. As we estimate more parameters (and decrease the
degrees of freedom), we are more likely to fit the data, but less likely to be
able to replicate the model with a second set of data. In other words, more
parameters may mean a loss of external validity. As we progress toward more
complex models (with increasingly complex parameterization, i.e. models 3 and 4
above), then we are more likely to fit the data better.

A number of notes regarding person-oriented analyses are important to raise at
the same time as their promise. Bauer (2007) notes that many samples of data can
be usefully broken down into profiles, and that the addition of profiles will
likely be suggested for reasons other than the samples coming from more than one
distribution (i.e., due to non-normality in the variables measured). Bauer also
cautions that profiles should not be reified; that profiles do not necessarily
exist outside of the analysis that they should be interpreted more as useful
interpretative devices. These cautions suggest that, in general, parsimony,
interpretability, and a general sense that the profiles are not necessarily
real, but are rather helpful analytic tools, should be both priorities for the
analyst and the reader of studies using this approach.

@@

SPSS is a common tool to carry out cluster analyses. SPSS provides functionality
to carry out both hierarchical and k-means cluster analysis. While somewhat
straightforward to carry out, particularly in SPSS’s graphical user interface
(GUI), there are some challenges to use of this approach. The GUI in SPSS can be
challenging, even for the most able analyst, to be able to document every step
with syntax, and so reproducing the entire analysis efficiently can be a
challenge, both for the analyst exploring various solutions and for the reviewer
looking to replicate this work. Additionally, SPSS is commercial software (and
is expensive), and so analysts without the software cannot carry out this
analysis.

Accordingly, we worked to develop tools, initially used by the research team and
now shared more widely, to provide basic functionality for each of the two
primary approaches described in this section, cluster analysis and Latent
Profile Analysis (LPA). In addition to being freely-available and open-source
(i.e., anyone can inspect the source code), the tools described lend themselves
to an open science. It is especially important to have an open transparent
methodology when using person-centered approaches because of the high dependence
on researcher judgment at multiple points in the process.

Accordingly, we set out to develop packages that a) provided sensible defaults
and were easy to use, but provided the option to access and modify all of the
inputs to the model (i.e., low barrier, high ceiling), b) interfaced to existing
tools,, and are able to translate between what existing tools are capable of and
what researchers and analysts carrying-out person-oriented analyses would like
to specify, c) made it easy to carry-out fully-reproducible analyses and d) were
well-documented. Both of these packages--one for two-step cluster analysis and
one for LPA--are designed to do one thing well, namely, quickly and easily
estimate profiles using cluster analysis or a model-based approach (Latent
Profile Analysis).

One way that tidyLPA is designed to be easy to use is that it assumes a “tidy”
data structure (Wickham, 2014). This means that it emphasizes the use of a data
frame as both the primary input and output of functions for the package. Because
data is passed to and returned (in amended form, i.e., with the latent profile
probabilities and classes appended to the data) from the function, it makes it
easy to create plots or use results in subsequent analyses. Another noteworthy
feature of tidyLPA is that it provides the same functionality through two
different tools, one that is open-source and available through R, the mclust
package (Scrucca et al., 2017) and one that is available through the commercial
software MPlus (Muthen & Muthen, 1997-2017). Moreover, as both tools use the
same maximum likelihood estimation procedure, they are benchmarked to produce
the same output (see here). Also, note that we have described the model
specifications with descriptions of what is estimated in terms of the variances
and covariances, the common names for the models (i.e., class-varying
unrestricted), and the covariance matrix associated with the parameterizationfor
the six models that are possible to be estimated on the website for tidyLPA (see
here).

# Analytic Choices

@@

## Whether to Use Cluster Analysis or LPA

Whether to use cluster analysis or LPA depends in part on the nature of the data
being investigated (cite the 2010 debate on this topic between Vermunt and
others; MacLachlan, 2011; Steinley & Brusco, 2011a; Steinley & Brusco, 2011b;
Vermunt, 2011). In some cases, cluster analysis and LPA solutions are very
similar (when a constrained LPA model is fit). Broadly, if the variables used
can be considered independent of one another, then cluster analysis may be
justified in terms of reaching an optimal solution. However, LPA provides other
benefits, such as the availability of information criteria (i.e., the Aikake
Information Criteria [AIC] and the Bayesian Information Criteria [BIC] and
statistical tests for the number of profiles (when a model is chosen, with a
bootstrapped likelihood-ratio test). LPA also results in probabilistic
assignments; this information can be used in subsequent analyses rather than the
simple assignment (Collins & Lanza, 2010).

## How to Choose the Number of Profiles

In the case of choosing the number of profiles (and the specification of the
model / profile solution), multiple criteria, such as the BIC or the proportion
of variance explained are recommended for decision-making, but also
interpretability in light of theory, parsimony, and evidence from
cross-validation should be considered. For cluster analysis, the proportion of
variance explained can be helpful for selecting a number of candidate profiles.
Then, the analyst should focus on the interpretability of the solution as well
as concerns of parsimony; if two profile solutions are similar, then the more
simple solution may be preferred.

## How to Choose Options for the Clustering Procedure or to Select the Model Parameterization

This applies most to LPA, but in cluster analysis, there are also options about
the distance metric used. In LPA, we can determine which parameters - variance
and covariance - are estimated to be the same or different across all profiles,
and, in the case of covariance, whether it is fixed to zero. Of course, for the
profiles we are most interested in, the mean is allowed to vary across profiles.
The models with more parameters freely-estimated use more degrees of freedom;
thus, similar to the number of profiles selected, the balance between adding
more parameters (and having a more complex model) and interpretability and
parsimony must be balanced. If a more complex model fits better and is
interpretable and is justifiable in terms of the data collected, then it may be
preferred to simpler models. If the two models are similar in terms of their
fit, then the more simple, parsimonious model should be selected.

# More General Recommendations

@@

# tidyLPA update

## Rationale for the update

**tidyLPA** has received a major update. Your old function calls might not work
as expected anymore. The new version of tidyLPA aims to make Latent Profile
Analysis (LPA) even more accessible to a broad audience, by means of a
user-friendly interface, and to fit more seemlessly into a 'tidyverse' workflow.

As before, **tidyLPA** offers a parallel workflow for the open-source package
Mclust, and the commercial package *Mplus*.

These changes necessitated some simplification of the existing functions.
Advanced options are still available, but require slightly more work on the
user's part. The upside of this is that, for most users, the tidyLPA workflow
and documentation are substantially simplified.

## Major changes

There have been major changes to the two primary functions,
`estimate_profiles()` and `compare_solutions()`

  - `estimate_profiles()` now estimates any number of profiles, for any number
    of models, using any algorithm. It has a simplified user interface. This
    makes it a more general function, of which the former version (which could
    only estimate one number of profiles for one model) is a special, restricted
    case.

  - `compare_solutions()` now compares solutions estimated by
    estimate\_profiles(), suggesting the optimal model and number of classes.
    This replaces its dual role of estimating AND comparing solutions. ` *
     `compare\_solutions()\` gives you the same set of fit indices for Mclust
    and Mplus. It provides a broader range of fit indices than before
      - `compare_solutions()` calls \`AHP()\`\`, a function which uses the
        Automatic Hierarchical Process to determine the optimal number of
        classes, based on several reputable fit indices

  - `plot_profiles()` now incorporates best practices for plotting latent
    profile analyses: Visualizing cluster centroids, standard errors, cluster
    variances, and classification uncertainty by means of weighted raw data

  - `impute_missing()` offers two easy-to-use single imputation options, for use
    with the Mclust package (which does not natively handle missing data). Mplus
    handles partially missing data by default.

The best way to understand these changes to **tidyLPA**, though, is by
explaining the new workflow.

## Example analysis using estimate\_profiles()

A full analysis might look like the following:

``` {r}
library(tidyverse)
library(tidyLPA)
```

pisaUSA15[1:100, ] %\>% select(broad\_interest, enjoyment, self\_efficacy)
%\>% single\_imputation() %\>% estimate\_profiles(3)

You can use Mplus simply by changing the package argument for
`estimate_profiles()`:

``` {r, eval = FALSE}
pisaUSA15[1:100, ] %>%
    select(broad_interest, enjoyment, self_efficacy) %>%
    single_imputation() %>%
    scale() %>%
    estimate_profiles(3) %>% 
    plot_profiles()
```

## Example analysis using compare\_solutions()

``` {r, eval = TRUE}
pisaUSA15[1:100, ] %>%
    select(broad_interest, enjoyment, self_efficacy) %>%
    single_imputation() %>%
    estimate_profiles(1:3, 
                      variances = c("equal", "varying"),
                      covariances = c("zero", "varying")) %>%
    compare_solutions(statistics = c("AIC", "BIC"))
```

## Deprecated functionality:

  - `compare_solutions()` no longer estimates profiles. It merely compares
    solutions estimated by `estimate_profiles()`

  - all Mplus-specific functions are deprecated. `estimate_profiles()` offers a
    unified interface to Mclust and Mplus. These Mplus-specific functions are
    deprecated:

    * `estimate_profiles_mplus()`
        * `compare_profiles_mplus()`
        * `plot_profiles_mplus()`

  - estimate\_profiles and compare\_solutions no longer select variables, passed
    as unquoted variable names, from 'df'. Instead, they use all variables in
    df. To select variables prior to analysis, use:

    * The dplyr function select(df, Your, Selected, Variable, Names)
        * The base R function df[, c("Your", "Selected", "Variable", "Names")]

  - estimate\_profiles and compare\_solutions no longer center and scale the raw
    data. To center and scale variables prior to analysis, use scale(df, center
    = TRUE, scale = TRUE)

  - estimate\_profiles no longer accepts arguments specific to Mplus or Mclust.
    Instead, you can specify additional arguments directly to the core functions
    `mplusModeler()` and `Mclust()`. See the help file for these functions for
    more details. Deprecated arguments, whose functionality is supplanted by
    passing arguments directly to these core functions, are:

For Mclust:

    * prior_control

For Mplus:

    * starts
    * m_iterations
    * st_iterations
    * convergence_criterion
    * optseed
    * cluster_ID
    * include_VLMR
    * include_BLRT (but BLRT is returned by default now)

# tidyLPA function with \_mplus() appended require Mplus to use

``` {r, eval = FALSE}
estimate_profiles_mplus(data, var1, var2, var3, ariances = “free”, covariances = “equal”) %>% 
    plot_profiles_mplus()
```

>>>>

# Introduction

@@

<!-- ## Cluster Analysis -->

<!-- Cluster analysis is an approach that has been frequently applied to analysis of -->
<!-- topics in educational psychology (such as motivation and engagement) from a more -->
<!-- holistic, whole-person approach. One particular way in which cluster analysis -->
<!-- has been used has been as part of an approach termed I-States-as-Objects -->
<!-- Analysis (ISOA). The purpose of ISOA is to identify profiles at specific time -->
<!-- points, and then to explore how individuals do or do not shift between profiles -->
<!-- across time points (needs cite). In their early work, Bergman and Magnusson -->
<!-- (1999) and colleagues (1999) described a two-step approach to cluster analysis -->
<!-- as part of their I-States-as-Objects analysis. In this approach, all of the -->
<!-- observations are clustered. The key idea behind the approach is that each -->
<!-- observation represents not a permanent group, but rather a status, or an -->
<!-- i-state. These i-states, are considered independently for the purposes of -->
<!-- constructing profiles. Then, shifts between these i-states across time (say, two -->
<!-- or three time points) are evaluated. -->

<!-- Commonly, cluster analysis carried out as part of ISOA in fact involves two -->
<!-- distinct clustering steps. This is because different clustering algorithms have -->
<!-- different strengths--as well as weaknesses. In ISOA, the first step is Ward’s -->
<!-- hierarchical cluster analysis (Bergman & El-Khouri, 1999). In this step, -->
<!-- observations are split, step-by-step, based upon their multivariate similarity. -->
<!-- The analyst provides the number of groups which will be used and interpreted, -->
<!-- though technically most algorithms will continue splitting the groups until each -->
<!-- observation makes up its own. This type of clustering is deterministic: The same -->
<!-- results will be obtained each time it is run. In the second step of the cluster -->
<!-- analysis, an iterative, k-means cluster analysis is carried out (Bergman and -->
<!-- El-Khouri, 1999). For this step, the algorithm partitions the observations into -->
<!-- groups, the success of the clustering based on a criterion such as the -->
<!-- variability in the data explained by the groups, and then “moves,” when -->
<!-- possible, the observations that are not in the group they are most similar to to -->
<!-- the one that they are most similar to. When no more such moves are possible, the -->
<!-- algorithm stops. On its own, k-means is not deterministic: Depending on the -->
<!-- starting values, it can obtain different results the different times that it is -->
<!-- run. So, the results from the hierarchical cluster analysis, the first step, are -->
<!-- used as start values to initialize the k-means clustering. -->

<!-- The rationale behind using these two approaches is that the hierarchical -->
<!-- clustering provides reasonable guesses for the profile membership, and then the -->
<!-- iterative clustering optimizes the clustering solution. To sum up, then, cluster -->
<!-- analysis as part of ISOA in fact involves two separate cluster analyses. The -->
<!-- merits of this approach are to leverage the benefits of each: the iterative -->
<!-- clustering starts with reasonable cluster solutions generated from the -->
<!-- hierarchical clustering and optimizes them. -->

# Latent Profile Analysis

The goal of LPA is estimate the parameters for a number of distributions
(typically multivariate) from a single dataset. Thus, such an approach is
model-based, and some descriptions in the literature refer to it as model-based
clustering (Hennig, Meila, Murtagh, & Rocci, 2015; Scrucca, Fop, Murphy, &
Raftery, 2017). Thus, one distinction between LPA and other, similar cluster
analytic approaches is that LPA is model-based; instead of using algorithms to
group together cases, LPA seeks to estimate parameters - in terms of variances
and covariances and how they are the same or different across profiles - that
best characterize the different distributions. Then, this approach seeks to
assign to each observation a probability that the observation is a sample from
the population associated with each profile (or mixture component).

Because LPA is model-based, a number of different model parameterizations can be
estimated. These models differ in terms of whether--and how--parameters are
estimated across the profiles. These parameters are the means for the different
profiles, which, in this approach, always are estimated freely across the
profiles; the variances for the variables used to create the profiles, which can
be estimated freely or can be estimated to be the same, or equal, across
profiles; and the covariances of the variables used to create the profiles,
which can be freely-estimated, estimated to be equal, or fixed to be zero.

We wish to note that challenge facing the analyst using LPA is that these
parameters and the distinct model parameterizations that can be estimated is the
different terminology used. As one example, Scrucca et al. (2017) refer to these
parameterizations not in terms of whether and how parameters are estimated, but
rather in terms of the geometric properties of the distributions that result
from particular parameterizations. Muthen and Muthen (1997-2017) and others
(Pastor et al., 2007) commonly refer to local independence to mean that the
covariances are fixed to zero (also described as the specification of the
covariance matrix as “diagonal,” because only the diagonal components, or the
variances, are estimated).

In general, as more parameters are estimated (i.e., those that are fixed to zero
are estimated as being equal across profiles; or those estimated as being equal
across profiles are freely-estimated across them), the model becomes more
complex; the model may fit better, but also be overfit, meaning that the
profiles identified may be challenging to replicate with another, separate data
set. Even still, flexibility in terms of which models can be estimated also has
affordances. For example, the varying means, equal variances, and covariances
fixed to 0. A researcher might choose this model specification if she wants to
model the variables to be used to create profiles that are independent. This
model is very simple, as no covariances are estimated and the variances are
estimated to be the same across profiles. As we estimate more parameters (and
decrease the degrees of freedom), we are more likely to fit the data, but less
likely to be able to replicate the model with a second set of data. In other
words, more parameters may mean a loss of external validity. As we progress
toward more complex models (with increasingly complex parameterization), then we
are more likely to fit the data better. More information on the model
parameterizations are discussed in the context of the software tool tidyLPA that
we have developed.

A note of caution is warranted about LPA in the context of their potential.
Bauer (2007) notes that many samples of data can be usefully broken down into
profiles, and that the addition of profiles will likely be suggested for reasons
other than the samples coming from more than one distribution (i.e., due to
non-normality in the variables measured). Bauer also cautions that profiles
should not be reified; that profiles do not necessarily exist outside of the
analysis that they should be interpreted more as useful interpretative devices.
These cautions suggest that, in general, parsimony, interpretability, and a
general sense that the profiles are not necessarily real, but are rather helpful
analytic tools, should be both priorities for the analyst and the reader of
studies using this approach.

@@

## Other Tools

SPSS is a common tool to carry out cluster analyses (but not LPA). While somewhat
straightforward to carry out, particularly in SPSS’s graphical user interface
(GUI), there are some challenges to use of this approach. The GUI in SPSS can be
challenging, even for the most able analyst, to be able to document every step
with syntax, and so reproducing the entire analysis efficiently can be a
challenge, both for the analyst exploring various solutions and for the reviewer
looking to replicate this work. Additionally, SPSS is commercial software (and
is expensive), and so analysts without the software cannot carry out this
analysis.

Another common tool is MPlus (Muthen & Muthen, 20xx). MPlus is a commercial tool
that provides functionality for many latent variable (and multi-level) models.
We will speak more to MPlus in the section on tidyLPA, as our software provides
an interface to both it and an open-source tool.

In R, a number of tools *can* be used to carry out LPA. OpenMx can be used for
this purpose (and to specify almost any model possible to specify within a
latent variable modeling approach). However, while OpenMx is very flexible, it
can also be challenging to use. Other tools in R allow for estimating Gaussian
mixture models, or models of multivariate Gaussian (or normal) distributions. In
this framework, the term “mixture component” has a similar meaning to a profile.
While much more constraining than the latent variable modeling framework, the
approach is often similar or the same: the EM algorithm is used to (aim to)
obtain the maximum likelihood estimates for the parameters being estimated. Like
in the latent variable modeling framework, different models can be specified.

In addition to following the same general approach, using tools that are
designed for Gaussian mixture modeling have other benefits, some
efficiency-related (see RMixMod, which uses compiled C++ code) and others in
terms of ease-of-use (i.e., the plot methods built-in to RMixMod, mclust, and
other tools). However, they also have some drawbacks, in that it can be
difficult to translate between the model specifications, which are often
described in terms of the geometric properties of the multivariate distributions
being estimated (i.e., “spherical, equal volume”), rather than in terms of
whether and how the means, variances, and covariances are estimated. They also
may use different default settings (than those encountered in MPlus) in terms of
the expectation-maximization algorithm, which can make comparing results across
tools challenging.

## tidyLPA

Because of the limitations in other tools, we set out to develop a tool that a) 
provided sensible defaults and were easy to use, but provided the option to access and modify all of the
inputs to the model (i.e., low barrier, high ceiling), b) interfaced to existing
tools,, and are able to translate between what existing tools are capable of and
what researchers and analysts carrying-out person-oriented analyses would like
to specify, c) made it easy to carry-out fully-reproducible analyses and d) were
well-documented.

This package focuses on models that are commonly specified as part of LPA.
Because MPlus is so widely-used, it can be helpful to compare output from other
software to MPlus. The functions in tidyLPA that use mclust have been
benchmarked to MPlus for a series of simple models (with small datasets and for
models with small numbers of profiles. This R Markdown output contains
information on how mclust and Mplus compare. The R Markdown to generate the
output is also available here, and, as long as you have purchased MPlus (and
installed MplusAutomation), can be used to replicate all of the results for the
benchmark. Note that most of the output is identical, though there are some
differences in the hundreths decimal places for some. Because of differences in
settings for the EM algorithm and particularly for the start values (random
starts for MPlus and starting values from hierarchical clustering for mclust),
differences may be expected for more complex data and models.

One way that tidyLPA is designed to be easy to use is that it assumes a “tidy”
data structure (Wickham, 2014). This means that it emphasizes the use of a data
frame as both the primary input and output of functions for the package. Because
data is passed to and returned (in amended form, i.e., with the latent profile
probabilities and classes appended to the data) from the function, it makes it
easy to create plots or use results in subsequent analyses. Another noteworthy
feature of tidyLPA is that it provides the same functionality through two
different tools, one that is open-source and available through R, the mclust
package (Scrucca et al., 2017) and one that is available through the commercial
software MPlus (Muthen & Muthen, 1997-2017). Moreover, as both tools use the
same maximum likelihood estimation procedure, they are benchmarked to produce
the same output. Also, note that we have described the model
specifications with descriptions of what is estimated in terms of the variances
and covariances, the common names for the models (i.e., class-varying
unrestricted), and the covariance matrix associated with the parameterizationfor
the six models that are possible to be estimated on the website for tidyLPA (see
here).

## Installation

You can install tidyLPA from CRAN with:

``` {r, eval = F}
install.packages("tidyLPA")
```

You can also install the development version of tidyLPA from GitHub with:

``` {r gh-installation, eval = FALSE}
install.packages("devtools")
devtools::install_github("data-edu/tidyLPA")
```

## Examples

<!-- This is where we need to use an example from our own data -->

### Mclust via tidyLPA

Here is a brief example using the built-in `pisaUSA15` data set and variables
for broad interest, enjoyment, and self-efficacy. Note that we first type the
name of the data frame, followed by the unquoted names of the variables used to
create the profiles. We also specify the number of profiles and the model. See
`?estimate_profiles` for more details.

In these examples, we pass the results of one function to the next by *piping*
(using the `%>%` operator, loaded from the `dplyr` package). We pass the data to
a function that selects relevant variables, and then to `estimate_profiles`:

``` {r, message = F}
library(tidyLPA)
library(dplyr)
```

``` {r}
pisaUSA15[1:100, ] %>%
    select(broad_interest, enjoyment, self_efficacy) %>%
    single_imputation() %>%
    estimate_profiles(3)
```

### Mplus via tidyLPA

We can use Mplus simply by changing the package argument for
`estimate_profiles()`:

### Plotting the profiles

``` {r}
pisaUSA15[1:100, ] %>%
    select(broad_interest, enjoyment, self_efficacy) %>%
    single_imputation() %>%
    scale() %>%
    estimate_profiles(3) %>% 
    plot_profiles()
```

## Model specification

In addition to the number of profiles (specified with the `n_profiles`
argument), the model can be specified in terms of whether and how the variable
variances and covariances are estimated.

The models are specified by passing arguments to the `variance` and `covariance`
arguments. The possible values for these arguments are:

  - `variances`: "equal" and "zero"
  - `covariances`: "varying", "equal", and "zero"

If no values are specified for these, then the variances are constrained to be
equal across classes, and covariances are fixed to 0 (conditional independence
of the indicators).

These arguments allow for four models to be specified:

  - Equal variances and covariances fixed to 0 (Model 1)
  - Varying variances and covariances fixed to 0 (Model 2)
  - Equal variances and equal covariances (Model 3)
  - Varying variances and varying covariances (Model 6)

Two additional models (Models 4 and 5) can be fit using MPlus. More information
on the models can be found in the
[vignette](https://data-edu.github.io/tidyLPA/articles/Introduction_to_tidyLPA.html).

Here is an example of specifying a model with varying variances and covariances
(Model 6; not run here):

``` {r, eval = F}
pisaUSA15[1:100, ] %>%
    select(broad_interest, enjoyment, self_efficacy) %>%
    single_imputation() %>%
    estimate_profiles(3, 
                      variances = "varying",
                      covariances = "varying")
```

## Comparing a wide range of solutions

``` {r, eval = FALSE}
pisaUSA15[1:100, ] %>%
    select(broad_interest, enjoyment, self_efficacy) %>%
    single_imputation() %>%
    estimate_profiles(1:3, 
                      variances = c("equal", "varying"),
                      covariances = c("zero", "varying")) %>%
    compare_solutions(statistics = c("AIC", "BIC"))
```

## Accessing values in the output

A few helper functions are available to make it easier to work with the output
of an analysis.

`get_data()` returns the data:

``` {r}
m <- pisaUSA15[1:100, ] %>%
    select(broad_interest, enjoyment, self_efficacy) %>%
    single_imputation() %>%
    estimate_profiles(3:4)
```

get\_data(m)

We note that `get_data()` returns data in wide format when applied to an object
of class tidyProfile (one element of a tidyLPA object), or when applied to a
tidyLPA object of length one. `get_data()` returns long format when applied to a
tidyLPA object containing multiple tidyProfile analyses (because then the wide
format does not make sense).

To transform data in the wide format into the long format, the `gather()`
function from the **tidyr** package can be used, e.g.:

``` {r}
get_data(m) %>% 
    tidyr::gather(Class_prob, Probability, contains("CPROB"))
```

# Recommendations Regarding Analytic Choices

@@

<!-- ## Whether to Use Cluster Analysis or LPA -->

<!-- Whether to use cluster analysis or LPA depends in part on the nature of the data -->
<!-- being investigated (cite the 2010 debate on this topic between Vermunt and -->
<!-- others; MacLachlan, 2011; Steinley & Brusco, 2011a; Steinley & Brusco, 2011b; -->
<!-- Vermunt, 2011). In some cases, cluster analysis and LPA solutions are very -->
<!-- similar (when a constrained LPA model is fit). Broadly, if the variables used -->
<!-- can be considered independent of one another, then cluster analysis may be -->
<!-- justified in terms of reaching an optimal solution. However, LPA provides other -->
<!-- benefits, such as the availability of information criteria (i.e., the Aikake -->
<!-- Information Criteria [AIC] and the Bayesian Information Criteria [BIC] and -->
<!-- statistical tests for the number of profiles (when a model is chosen, with a -->
<!-- bootstrapped likelihood-ratio test). LPA also results in probabilistic -->
<!-- assignments; this information can be used in subsequent analyses rather than the -->
<!-- simple assignment (Collins & Lanza, 2010). -->

## How to Choose the Number of Profiles/Classes

In the case of choosing the number of profiles (and the specification of the
model / profile solution), multiple criteria, such as the BIC or the proportion
of variance explained are recommended for decision-making, but also
interpretability in light of theory, parsimony, and evidence from
cross-validation should be considered.

<!-- ## How to Choose Options for the Clustering Procedure or to Select the Model Parameterization -->

<!-- This applies most to LPA, but in cluster analysis, there are also options about -->
<!-- the distance metric used. In LPA, we can determine which parameters - variance -->
<!-- and covariance - are estimated to be the same or different across all profiles, -->
<!-- and, in the case of covariance, whether it is fixed to zero. Of course, for the -->
<!-- profiles we are most interested in, the mean is allowed to vary across profiles. -->
<!-- The models with more parameters freely-estimated use more degrees of freedom; -->
<!-- thus, similar to the number of profiles selected, the balance between adding -->
<!-- more parameters (and having a more complex model) and interpretability and -->
<!-- parsimony must be balanced. If a more complex model fits better and is -->
<!-- interpretable and is justifiable in terms of the data collected, then it may be -->
<!-- preferred to simpler models. If the two models are similar in terms of their -->
<!-- fit, then the more simple, parsimonious model should be selected. -->

## More General Recommendations

@@
